Dialogue act classification is a laughing matter

### 2021-09-19 reports/SemDial2021/semdial.tex
merge SemDial final version

%\aclfinalcopy % Uncomment this line for the final submission
%\def\aclpaperid{***} %  Enter the acl Paper ID here

@@

\author{First Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  \texttt{email@domain} \\\And
  Second Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  \texttt{email@domain} \\}

@@

Laughter is ubiquitous in our everyday interactions.  In the
Switchboard Dialogue Act Corpus \citep[SWDA,][]{jurafsky1997switchboard} (US
English, phone conversations where two participants that
are not familiar with each other discuss a potentially controversial
subject, such as gun control or the school system) non-verbally vocalised
dialogue acts (whole utterances that are marked as non-verbal, 66\% of
which contain laughter) constitute 1.7\% of all dialogue acts.
Laughter tokens\footnote{Switchboard Dialogue Act Corpus does not include speech-laughs.} make up 0.5\% of all the tokens that occur in the
corpus.

Laughter relates to the discourse structure of dialogue
and can refer to a \emph{laughable}, which can be a perceived event or an
entity in the discourse.
Laughter can precede, follow or overlap the laughable, with the time
alignment between the laughter and laughable dependent on who produces
the laughable, the form of the laughter, and the pragmatic function performed \citep{tian2016we}. 
%Laughter can act as a contextual feature for determining the sincerity of an utterance, e.g. when detecting sarcasm
% \citep{tepperman2006yeah}.% ??
% VM (removed as too vague!): Some studies have already pointed out the role of laughter to disambiguate between literal or not literal interpretations of utterances.

@@

In this paper we investigate the disambiguation potential of laughter
for interpreting the meaning of an utterance, with the central point
being the dialogue act performed by it. To do so, we employ a
Transformer-based model and look into laughter as a potentially useful
feature for the task of \emph{dialogue act recognition} (DAR). Laughs are not
present in a large-scale pre-trained models, such as BERT \citep{Devlin2019a}, but their
representations can be learned while training for a dialogue-specific
task (DAR in our case). We further explore whether such representations
can be additionally learned, in an unsupervised fashion, from
dialogue-like data, such as a movie subtitles corpus, and if it
further improves the performance of our model. 
% Vlad check if it is right what I have written!! XD XD
% VM: perfect

@@

Moreover \citet{romaniuk2009clinton} and \citet{ginzburg2020laughter}
discuss how laughter can answer or decline to answer a question, and
\citet{mazzocconi2018laughter} discusses how even isolated laughter
can be the object of clarification requests, signalling the need for
interlocutors to clarify its meaning (e.g., in terms of what the
``laughable'' is) in order to continue the conversation.

@@

Let us illustrate our comparison schema using the other two outliers,
\emph{Downplayer} (make up 0.05\% of all utterances) and \emph{Apology} (0.04\%), comparing them with the most common dialogue
act in SWDA---\emph{Statement-Non-Opinion} (33.27\%).  We take into consideration five
laughter-related dimensions of an utterance, and create 5-dimensional
(pentagonal) representations of DAs according to them. Each
dimension's value is equal to the proportion of utterances of a given
type which contain laughter:
\begin{itemize}
\item[$\uparrow$] current utterance;
\item[$\nwarrow$] immediately preceding utterance by the same speaker;
\item[$\nearrow$] immediately following utterance by the same speaker;
\item[$\swarrow$] immediately preceding utterance by the other speaker;
\item[$\searrow$] immediately following utterance by the other speaker.
\end{itemize}

@@

\begin{lingex}%CH What are # ?
\item\label{ex:apology-downplayer}\footnote{Overlapping material is marked wish hash signs.}
  \small
  \adjustbox{valign=t}{
    % \setlength\tabcolsep{1.5pt}
    \begin{tabulary}{\linewidth}{@{}>{\bf}lJ>{\em}r@{}}
      A: & I'm sorry to keep you waiting \#<laughter>.\# & Apology \\
      B: & \#Okay\# <laughter>. / & Downplayer \\
      A: & Uh, I was calling from work  & Statement (n/o) \\
\end{tabulary}}
\end{lingex}

@@

\subsection{Observations}
\label{sec:observations}
\paragraph{Laughter and modification or enrichment of the current DA}
We observe a higher proportion of laughter accompanying the current
dialogue act ($\uparrow$) when the laughter is aimed at modifying
the current dialogue act with some degree of urgency in order to
smooth or soften it (\emph{Action-directive}, \emph{Reject},
\emph{Dispreferred answer}, \emph{Apology}), to contribute to its
enrichment stressing the positive disposition towards the partner
(\emph{Appreciation}, \emph{Downplayer}, \emph{Thanking}), or to cue
for the need to consider a less probable meaning, therefore helping
in non-literal meaning interpretation (\emph{Rhetorical question}).

@@

A peculiar case is than the one of \emph{Self-talk}, often followed by
laughter by the same speaker. In this case the laughter may be
produced to signal the incongruity of the action (in dialogue we
normally speak to others, not to ourselves), while at the same time
function to smooth the situation, for instance, when having issues of
lexical retrieval, as in \ref{ex:selftalk}, or some degree of
embarrassment from the speaker, when questioning whether a
contribution is appropriate or not, as in
\ref{ex:appropriate}.

@@

The interesting mirror-image %specular\todo{does it work in english? BN: I didn't know this word before but I think so? CH: I have no idea what it means. Can we rephrase so it's understandable to mere mortals??}
patterns
observable in the lower part of the graph can therefore be explained
by considering the relation between the two dialogues acts.  We
observe cases in which an \textit{Apology} is accompanied by a
laughter, and then followed by a \textit{Downplayer}, showing 
that the laughter’s positive effect was attained and successful. This
allows us to explain both the bottom left spike ($\swarrow$) observed for
\textit{Downplayer} (often preceded by an utterance by the
partner containing laughter) and the bottom right spike ($\searrow$) observed
for \emph{Apology} (often followed by an utterance by the
partner containing laughter). In example \ref{ex:apology-downplayer}
both the apology and the downplayer are accompanied by laughter, while
in \ref{ex:sorry} a typical example of a laughter accompanying an \emph{Apology}
is followed by a \emph{Downplayer}.

@@

We now turn to the question of whether our qualitative observations of patters between laughs and dialogue acts can be used to improve a dialogue act recognition task.

@@

As a baseline utterance encoder, we use a word-level CNN with window sizes of 3, 4, and 5, each with 100 feature maps \citep{kimConvolutionalNeuralNetworks2014}. 
The model uses 100-dimensional word embeddings, which are initialised with pre-trained gloVe vectors \citep{penningtonGloveGlobalVectors2014}.

For the BERT utterance encoder, we use the BERT\textsubscript{BASE} model with hidden size of 768 and 12 transformer layers and self-attention heads \citep[][\S3.1]{devlinBERTPretrainingDeep2018}.
In our implementation, we use the un-cased model provided by \citet{wolfHuggingFaceTransformersStateoftheart2019}.

The next section will introduce the experiments that we have carried out and will discuss the results that we obtained from them.

@@

First of all, we can see that  \texttt{BERT}  outperforms \texttt{CNN} in all cases.
From observing the effect of laughters, we can see differences in performance depending on the dialogue act, regardless of how often laughter occurs in the current or adjacent utterances (see Figure~\ref{fig:by-da} in Appendix~\ref{sec:suppl}).
The strongest evidence for laughter as a helpful feature was found in SWDA for the \texttt{BERT} utterance encoder, where macro-F1 score increases by 7.89 percentage points.

@@

  \begin{figure}
  \centering
    \fontsize{6}{10}\selectfont
%      \includesvg[width=\linewidth]{img/da-svd-verbal}
   \includegraphics[width=\linewidth]{img/swda-cm.pdf}
  \caption{Confusion matrices for \texttt{BERT-NL} (top) vs \texttt{BERT-L} (bottom); SWDA corpus. Solid lines show classification improvement  of rhetorical questions.}
    \label{fig:swda-cm}
  \end{figure}

@@

As previously noted, training data for BERT does not include
features specific to dialogue (e.g.\ laughs). We therefore  experiment
with a large and more dialogue-like corpus constructed from
OpenSubtitles \citep{Lison2016} (350M tokens, where 0.3\% are laughter
tokens).  We used a manually constructed list of words frequently used
to refer to laughter in subtitles and replaced every occurance of one
of these words with the special laughter token.  We then collected
every English-language subtitle file in which at least 1\% of the
utterances contained laughter (about 11\% of the total).  Because
utterances are not labelled with speaker in the OpenSubtitles corpus,
we randomly assigned a speaker token to each utterance to maintain the
format of the other dialouge corpora.

@@

For 314 (76\%) of such dialogue acts the model has predicted the
\emph{Acknowledge (Backchannel)} class and for 46 (11\%) ---
continuations of the previous DA by the same speaker. The rest were
classified as either something uninformative (the \emph{Abandoned or
  Turn-Exit or Uninterpretable} class) or, from manual observation,
clearly unrelated.

\emph{Acknowledge (Backchannel)} can cover some uses of laughter, for
instance, to show to the interlocutor acknowledgement of their
contribution, implying the appreciation of an incongruity, and
inviting continuation (functioning simultaneously as a continuer and
assessment feedback
\citep{schegloff1982discourse}), as in example~\ref{ex:fish}.

@@

\section{Discussion}
\label{sec:discussion}
The implications of the results obtained are twofold: showing that
laughter can help a computational model to attribute meaning to an
utterance and help with pragmatic disambiguation, and consequently
stressing once again the need for integrating laughter (and other
non-verbal social signals) in any framework aimed to model meaning in
interaction \citep{ginzburg2020laughter,Maraev.Mazzocconi.Howes.Ginzburg_ISCA_2018}. % TODO: and, in final version add LACATODA

@@

Our results provide further evidence
(e.g. \citet{torres1997modeling,gazelaughter}) for the fact that
non-verbal behaviours are tightly related to the dialogue information
structure, propositional content and  dialogue act performed by
utterances.
Laughter, along with other non-verbal social signals, can constitute a
dialogue act in itself conveying meaning and affecting the unfolding
 dialogue \citep{bavelas2000visible,ginzburg2020laughter}.

@@

There is abundant room for further progress in determining how other speech-related information, such as prosody and disfluencies, can be incorporated into a DAR model.
\citet{stolckeDialogueActModeling2000} showed that dialogue acts can have specific prosodic manifestations that can be used to improve DAR.
Furthermore, in relation to laughter, its form (duration, arousal, overlap with speech) can be informative about its function and position w.r.t.\ the laughable \citep{tian2016we,mazzocconi2019phd}.
Incorporating such information is crucial if models pre-trained on large-scale text corpora are to be adapted for use in dialogue applications.

@@

% \section*{Acknowledgments}

>>>>

\aclfinalcopy % Uncomment this line for the final submission
%\def\aclpaperid{***} %  Enter the acl Paper ID here

@@

\author{Vladislav Maraev \\
  University of Gothenburg \\
  \texttt{vladislav.maraev@gu.se} \\\And
  Bill Noble \\
  University of Gothenburg \\
  \texttt{bill.noble@gu.se} \\\AND
  Chiara Mazzocconi  \\
  Aix-Marseille University \\
  \texttt{chiara.mazzocconi@live.it} \\\And
  Christine Howes \\
  University of Gothenburg \\
  \texttt{christine.howes@gu.se} \\}

@@

Laughter is ubiquitous in our everyday interactions.  In the
Switchboard Dialogue Act Corpus \citep[SWDA,][]{jurafsky1997switchboard} (US
English, phone conversations where two participants that
are not familiar with each other discuss a potentially controversial
subject, such as gun control or the school system) non-verbally vocalised
dialogue acts (whole utterances that are marked as non-verbal, 66\% of
which contain laughter) constitute 1.7\% of all dialogue acts.
Laughter tokens\footnote{Switchboard Dialogue Act Corpus does not include speech-laughs.} make up 0.5\% of all the tokens that occur in the
corpus.
%
Laughter relates to the discourse structure of dialogue
and can refer to a \emph{laughable}, which can be a perceived event or an
entity in the discourse.
Laughter can precede, follow or overlap the laughable, and the time alignment between them depends on who produces
the laughable, the form of the laughter, and the pragmatic function performed \citep{tian2016we}. 
%Laughter can act as a contextual feature for determining the sincerity of an utterance, e.g. when detecting sarcasm
% \citep{tepperman2006yeah}.% ??
% VM (removed as too vague!): Some studies have already pointed out the role of laughter to disambiguate between literal or not literal interpretations of utterances.

@@

Based on previous work and the corpus study presented in this paper, we argue that laughter is tightly related to the information structure of a dialogue. In this paper, we investigate the potential of laughter
to disambiguate the meaning of an utterance, in terms of 
the dialogue act it performs. To do so, we employ a
Transformer-based model and look into laughter as a potentially useful
feature for the task of \emph{dialogue act recognition} (DAR). Laughs are not
present in a large-scale pre-trained models, such as BERT \citep{Devlin2019a}, but their
representations can be learned while training for a dialogue-specific
task (DAR in our case). We further explore whether such representations
can be additionally learned, in an unsupervised fashion, from
dialogue-like data, such as a movie subtitles corpus, and if it
further improves the performance of our model. 
% Vlad check if it is right what I have written!! XD XD
% VM: perfect

@@

Moreover \citet{romaniuk2009clinton} and \citet{ginzburg2020laughter}
discuss how laughter can answer or decline to answer a question, and
\citet{mazzocconi2018laughter} explore  laughter
as an object of clarification requests, signalling the need for
interlocutors to clarify its meaning (e.g., in terms of what the
``laughable'' is) to carry on with the conversation.

@@

Let us illustrate our comparison schema using the other two outliers,
\emph{Downplayer} (make up 0.05\% of all utterances) and \emph{Apology} (0.04\%), comparing them with the most common dialogue
act in SWDA -- \emph{Statement-Non-Opinion} (33.27\%).  We consider
laughter-related dimensions of an utterance, and create 5-dimensional
(pentagonal) representations of DAs according to them. Each
dimension's value is equal to the proportion of utterances of a given
type which contain laughter:
\begin{itemize}
\setlength\itemsep{0em}
\item[$\uparrow$] current utterance;
\item[$\nwarrow$] immediately preceding utterance by the same speaker;
\item[$\nearrow$] immediately following utterance by the same speaker;
\item[$\swarrow$] immediately preceding utterance by the other speaker;
\item[$\searrow$] immediately following utterance by the other speaker.
\end{itemize}

@@

\begin{lingex}%CH What are # ?
\item\label{ex:apology-downplayer}\footnote{Overlapping material is marked with hash signs.}
  \small
  \adjustbox{valign=t}{
    % \setlength\tabcolsep{1.5pt}
    \begin{tabulary}{\linewidth}{@{}>{\bf}lJ>{\em}r@{}}
      A: & I'm sorry to keep you waiting \#<laughter>.\# & Apology \\
      B: & \#Okay\# <laughter>. / & Downplayer \\
      A: & Uh, I was calling from work  & Statement (n/o) \\
\end{tabulary}}
\end{lingex}

@@

\subsection{Observations}
\label{sec:observations}
\paragraph{Laughter and modification or enrichment of the current DA}
We observe a higher proportion of laughter accompanying the current
dialogue act ($\uparrow$) when the laughter is aimed at modifying
the current dialogue act with some degree of urgency to
smooth or soften it (\emph{Action-directive}, \emph{Reject},
\emph{Dispreferred answer}, \emph{Apology}), to contribute to its
enrichment stressing the positive disposition towards the partner
(\emph{Appreciation}, \emph{Downplayer}, \emph{Thanking}), or to cue
for the need to consider a less probable meaning, therefore helping
in non-literal meaning interpretation (\emph{Rhetorical question}).

@@

A peculiar case is the one of \emph{Self-talk}, often followed by
laughter by the same speaker. In this case the laughter may be
produced to signal the incongruity of the action (in dialogue we
normally speak to others, not to ourselves), while at the same time
function to smooth the situation, for instance, when having issues of
lexical retrieval, as in \ref{ex:selftalk}, or some degree of
embarrassment from the speaker, when questioning whether a
contribution is appropriate or not, as in
\ref{ex:appropriate}.

@@

The interesting mirror-image %specular\todo{does it work in english? BN: I didn't know this word before but I think so? CH: I have no idea what it means. Can we rephrase so it's understandable to mere mortals??}
patterns
observable in the lower part of the graph can therefore be explained
by considering the relation between the two dialogue acts.  We
observe cases in which an \textit{Apology} is accompanied by a
laughter, and then followed by a \textit{Downplayer}, showing 
that the laughter’s positive effect was attained and successful. This
allows us to explain both the bottom left spike ($\swarrow$) observed for
\textit{Downplayer} (often preceded by an utterance by the
partner containing laughter) and the bottom right spike ($\searrow$) observed
for \emph{Apology} (often followed by an utterance by the
partner containing laughter). In example \ref{ex:apology-downplayer}
both the apology and the downplayer are accompanied by laughter, while
in \ref{ex:sorry} a typical example of a laughter accompanying an \emph{Apology}
is followed by a \emph{Downplayer}.

@@

We now turn to the question of whether our qualitative observations of patterns between laughs and dialogue acts can be used to improve a dialogue act recognition task.

@@

As a baseline utterance encoder, we use a word-level CNN with window sizes of 3, 4, and 5, each with 100 feature maps \citep{kimConvolutionalNeuralNetworks2014}. 
The model uses 100-dimensional word embeddings, which are initialised with pre-trained gloVe vectors \citep{penningtonGloveGlobalVectors2014}.
%
For the BERT utterance encoder, we use the BERT\textsubscript{BASE} model with hidden size of 768 and 12 transformer layers and self-attention heads \citep[][\S3.1]{devlinBERTPretrainingDeep2018}.
In our implementation, we use the un-cased model provided by \citet{wolfHuggingFaceTransformersStateoftheart2019}.

% The next section will introduce the experiments that we have carried out and will discuss the results that we obtained from them.

@@

\texttt{BERT}  outperforms the \texttt{CNN} on AMI-DA. On SWDA, the two encoders are more comparable, though \texttt{BERT} has a slight edge in accuracy, suggesting that it relies more heavily on defaulting to common dialogue act tags.
On SWDA, we see small improvements in accuracy and macro-F1 for models that included laughter. 
For AMI-DA, the effect of laughter is small or even negative -- the impact of laughter on performance becomes more clear in the disaggregated performance over different dialogue acts. 
Indeed, laughter improves the accuracy of the model even on some dialogue acts in which laughter occurs rarely in the current and adjacent utterances (see Figure~\ref{fig:by-da} in Appendix~\ref{sec:suppl}).

@@

  \begin{figure}[h!]
  \centering
  \vspace*{-0.3cm}
    \fontsize{6}{10}\selectfont
%      \includesvg[width=\linewidth]{img/da-svd-verbal}
   \includegraphics[width=.9\linewidth]{img/swda-cm.pdf}
  \caption{Confusion matrices for \texttt{BERT-NL} (top) vs \texttt{BERT-L} (bottom); SWDA corpus. Solid lines show classification improvement  of rhetorical questions.}
%  \vspace*{-0.7cm}
    \label{fig:swda-cm}
  \end{figure}

@@

As previously noted, training data for BERT does not include
features specific to dialogue (e.g.\ laughs). We therefore  experiment
with a large and more dialogue-like corpus constructed from
OpenSubtitles \citep{Lison2016} (350M tokens, where 0.3\% are laughter
tokens).  We used a manually constructed list of words frequently used
to refer to laughter in subtitles and replaced every occurance of one
of these words with the special laughter token.  We then collected
every English-language subtitle file in which at least 1\% of the
utterances contained laughter (about 11\% of the total).  Because
utterances are not labelled with speaker in the OpenSubtitles corpus,
we randomly assigned a speaker token to each utterance to maintain the
format of the other dialogue corpora.

@@

For 314 (76\%) of such dialogue acts the model has predicted the
\emph{Acknowledge (Backchannel)} class and for 46 (11\%) --
continuations of the previous DA by the same speaker. The rest were
classified as either something uninformative (the \emph{Abandoned or
  Turn-Exit or Uninterpretable} class) or, from manual observation,
clearly unrelated.

\emph{Acknowledge (Backchannel)} can cover some uses of laughter, for
instance, to show to the interlocutor acknowledgement of their
contribution, implying the appreciation of an incongruity and
inviting continuation, functioning simultaneously as a continuer and
assessment feedback
\citep{schegloff1982discourse}, as in example~\ref{ex:fish}.

@@

\section{Discussion}
\label{sec:discussion}
The implications of the results obtained are twofold: showing that
laughter can help a computational model to attribute meaning to an
utterance and help with pragmatic disambiguation, and consequently
stressing once again the need for integrating laughter (and other
non-verbal social signals) in any framework aimed to model meaning in
interaction \citep{ginzburg2020laughter,maraev2021lacatoda}. % TODO: and, in final version add LACATODA

@@

Our results provide further evidence
(e.g. \citet{torres1997modeling,mazzocconi2021looking}) for the fact that
non-verbal behaviours are tightly related to the dialogue information
structure, propositional content and  dialogue act performed by
utterances.
Laughter, along with other non-verbal social signals, can constitute a
dialogue act in itself conveying meaning and affecting the unfolding
 dialogue \citep{bavelas2000visible,ginzburg2020laughter}.

@@

% There is abundant room for further progress in determining how other speech-related information, such as prosody and disfluencies, can be incorporated into a DAR model.
% \citet{stolckeDialogueActModeling2000} showed that dialogue acts can have specific prosodic manifestations that can be used to improve DAR.
% Furthermore, in relation to laughter, its form (duration, arousal, overlap with speech) can be informative about its function and position w.r.t.\ the laughable \citep{mazzocconi2019phd}.
% Incorporating such information is crucial if models pre-trained on large-scale text corpora are to be adapted for use in dialogue applications.

There is abundant room for further study of how laughter can help to disambiguate communicative intent. 
\citet{stolckeDialogueActModeling2000} showed that the specific prosodic manifestations of an utterance can be used to improve DAR.
With respect to laughter, the form (duration, arousal, overlap with speech) can be informative about its function and position w.r.t.\ the laughable \citep{mazzocconi2019phd}.
Incorporating such information is crucial if models pre-trained on large-scale text corpora are to be adapted for use in dialogue applications.

@@

\section*{Acknowledgments}
Maraev, Noble and Howes were supported by  the Swedish Research Council (VR) grant  2014-39 for the establishment of the Centre for Linguistic Theory and Studies in Probability (CLASP). Mazzocconi was supported by the “Investissements d’Avenir” French government program managed by the French National Research Agency (reference: ANR-16-CONV-0002) and from the Excellence Initiative of Aix-Marseille University—“A*MIDEX” through the Institute of Language, Communication and the Brain.
